PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-04-19 papers/NAR_web_server_2009/paper.tex
Modifications to resubmission

\history{Received Feb. 13, 2010}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \mc{(version
  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
\cite{Huang99} (the latter of which implicitly calculates a clustering
in the process of assembly).  For our simulation tests we used the
{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
fragments of varying length under different models of error
(Supplementary Materials, Section~D.1), generating the fragments from
the list of 100 zebra fish genes used in the \wcd\/ testing
\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
as possible, those used in the \wcd\/ study (see Supplementary
Materials).  \mc{The most important method of quality assessment is
  {\it sensitivity} (the fraction of fragment pairs from the same gene
  that were correctly clustered together).  But we also look at the
  {\it Jaccard Index} (which balances sensitivity with the number of
  false positives by looking at the ratio of true positives to the sum
  of true positives, false positives, and true negitives), {\it Type 1
    error} (the fraction of genes that were divided between clusters),
  and {\it Type 2 error} (the fraction of clusters containing two or
  more genes).  Specificity was not an issue in simulated tests, as
  both \peace\/ and \wcd\/ were completely successful in separating
  unrelated ESTs (save when dealing with recently duplicated genes).
  Other measurements are also discussed in the supplementary
  materials.  In Figure~\ref{SeJiT1T2} we plot these four tests as a
  function of erro rate, and observe the almost identical results
  between \peace\/ and \wcd\/.  In Figure~\ref{seq_parallel} we plot
  the runtime for \peace\/ and \wcd\/, again observing almost identical
  results when run sequentially -- but signifcantly faster runtime for
  PEACE on multiple processer when holding the EST/processor ratio
  constant (ranging from a 65\% improvement for two processors too a
  17\% improvement for 12 processors).}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
  \caption{Comparisons of Sensitivity, Jiccard Index, Type 1 error and
    Type 1 error, based on the
    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
    fish genes  (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
    tics = 95\% confidence intervals on estimates.  Intervals are not
    presented for Type 1 error due to the effective lack of
    variance.}\label{SeJiT1T2}
\end{figure}

@@

\mc{In applying the tools to real data Sanger data, we used the Human
Benchmark Dataset used to test \easycluster\/ \cite{Picardi09}, and
the A076941.fa {\it Arabidopsis thaliana} dataset used to test WCD
\cite{Hazelhurst08,Hazelhurst08a}, with results in
Table~\ref{SangerReal}.  We notice essentially identical
results for PEACE and WCD in quality, both significantly better than
Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
with PEACE showing a 60\% runtime improvement in the smaller dataset,
requiring 20\% more time in the Arabidopsis dataset.}

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\label{SangerReal}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Dataset and the WCD A076941 Arabidopsis thaliana dataset using the
  standard quality measurments.}
\end{table*}

@@

We were unable to run \wcd\/ on short-read sequences, as the
distributed implementation is unable to handle such data without an
adjustment to the code.  Hence, while \peace\/ was run on both
simulated and real NGS fragment sets, there was no cluster tool to
which the results could be compared.  \peace\/ did well, achieving a
sensitivity of 0.75 when applied to a simulated short-read set
subjected to a sequencing error rate of 2\% (with comparable {\it
  Jaccard Index} values reflecting both sensitivity {\it and}
specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
approximately 17.6 minutes to process the 7.9 Mb of data when run in
sequential mode on a .  When applied to a 64 Mb set of
165,490 {\it
 Chlamydomonas reinhardtii} transcript fragments
sequenced by 454 pyrosequencing technology (ranging in size from 50
bases to several hundred), \peace\/ completed the job in 49 minutes
using 30 of the Xeon processors.

@@

% LocalWords:  arallel nalysis lustering ngine Rao Moler Ozden Zhang Liang ESTs
% LocalWords:  Karro XXXXX transcriptome pre isoforms SNPs unsequenced runtime
% LocalWords:  chlamydomondomonas ESTate PaCE xsact CLU WCD Hazelhurst et al th
% LocalWords:  CHUN nalyzer Prim's MSTs subgraph MPI SourceCode ESTSim runtimes
% LocalWords:  ests Chlamydomonas Reihartdii gmap CDNA transcriptomes cDNA www
% LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo fasta
% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots nvironment
% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
% LocalWords:  NGS polyA set's GHz Xeon Bikun

>>>>

\history{Received Feb. 13, 2010 \\ Resubmitted April 20, 2010}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \mc{(version
  0.5.1)}\cite{Hazelhurst08a} and the \capthree\/ assembly tool
\cite{Huang99} (the latter of which implicitly calculates a clustering
in the process of assembly).  For our simulation tests we used the
{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
fragments of varying length under different models of error
(Supplementary Materials, Section~D.1), generating the fragments from
the list of 100 zebra fish genes used in the \wcd\/ testing
\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
as possible, those used in the \wcd\/ study (see Supplementary
Materials).  \mc{The most important method of quality assessment is
  {\it sensitivity} (the fraction of fragment pairs from the same gene
  that were correctly clustered together).  But we also look at the
  {\it Jaccard Index} (which balances sensitivity with the number of
  false positives by looking at the ratio of true positives to the sum
  of true positives, false positives, and true negatives), {\it Type 1
    error} (the fraction of genes that were divided between clusters),
  and {\it Type 2 error} (the fraction of clusters containing two or
  more genes).  Specificity was not an issue in simulated tests, as
  both \peace\/ and \wcd\/ were completely successful in separating
  unrelated ESTs (save when dealing with recently duplicated genes).
  Other measurements are also discussed in the supplementary
  materials.  In Figure~\ref{SeJiT1T2} we plot these four tests as a
  function of error rate, and observe the almost identical results
  between \peace\/ and \wcd\/.  In Figure~\ref{seq_parallel} we plot
  the runtime for \peace\/ and \wcd\/, again observing almost identical
  results when run sequentially -- but significantly faster runtime for
  PEACE on multiple processor when holding the EST/processor ratio
  constant (ranging from a 65\% improvement for two processors too a
  17\% improvement for 12 processors).}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.pdf}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the
    average over 30 simulated Sanger Sequence ESTs sets derived from 100 zebra
    fish genes  (see
    Supplementary Materials, Section C, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd, Black/Dot-Dash = \capthree; vertical
    tics = 95\% confidence intervals on estimates.  Intervals are not
    presented for Type 1 error due to the effective lack of
    variance.}\label{SeJiT1T2}
\end{figure}

@@

\mc{In applying the tools to real data Sanger data, we used the Human
Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
the A076941.fa {\it Arabidopsis thaliana} data set used to test WCD
\cite{Hazelhurst08a,Hazelhurst08b}, with results in
Table~\ref{SangerReal}.  We notice essentially identical
results for PEACE and WCD in quality, both significantly better than
Cap3 in Sensitivity and Type 1 error rate, while slightly worse in the
Jaccard Index and Type 2 error.  In runtime we see some inconsistency,
with PEACE showing a 60\% runtime improvement in the smaller data set,
requiring 20\% more time in the Arabidopsis data set.  In Section~D.1
of the Supplementary Materials we present runtimes several more
sets, observing that while \peace\/ appears to be significantly more
faster on the smaller sets, \wcd\/ does overtake it for larger sets.}

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline Bicyclist Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\label{SangerReal}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}
\end{table*}

@@

\mc{We tested the three tools on short-read data using the MetaSim
  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
  MetaSim are sequence generation and error models corresponding to
  several technologies, including the 454 short-read technology.
  Leaving \capthree\/ and \metasim\/ at their default values, but
  employing the {\it adaptive $d^2$} strategy for short read sequences
  (see Section~C in the supplementary materials), we find a that
  \peace\/ performs significantly better than the other two tools on
  454 data using the default MetaSim model (with sequence sizes
  ranging in size from 200bp to 370bp, at an average size of 250bp and
  a standard deviation of 17bp).  Over thirty runs, we see an average
  sensitivity 0.884 for \peace, 0.331\/ for \wcd, and 0.004 for
  \capthree.  For the Jaccard Index we get essentially the same
  numbers.  For 100 transcript \peace creates 128 clusters (as opposed
  to 6,370 and 19245 clusters for \wcd and \capthree), and creates
  only 3 singletons (as opposed to 5,850 and 15432).  We find it
  likely that the \wcd results could be considerably improved by the
  simple adjustment of a few parameters, but were unable to determine
  how to do that.  (Runtime of \peace\/ was somewhat slower than that
  of \wcd, but given the significant improvement in quality results we
  do not find this to be very troubling.)}

@@

% LocalWords:  arallel nalysis lustering ngine Rao Moler Ozden Zhang Liang ESTs
% LocalWords:  Karro XXXXX transcriptome pre isoforms SNPs unsequenced runtime
% LocalWords:  chlamydomondomonas ESTate PaCE xsact CLU WCD Hazelhurst et al th
% LocalWords:  CHUN nalyzer Prim's MSTs subgraph MPI SourceCode ESTSim runtimes
% LocalWords:  ests Chlamydomonas Reihartdii gmap CDNA transcriptomes cDNA www
% LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo fasta
% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots nvironment
% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV
% LocalWords:  Data Set thaliana data set EasyCluster MetaSim
